COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-23 TACO-paper/coffee.tex
went or adjusting stuff

\usepackage{tabularx}
\usepackage{tabulary}
\usepackage[cmex10]{amsmath}
\usepackage{url}

@@

% Title portion
\title{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
\author{GANG ZHOU
\affil{College of William and Mary}
YAFENG WU
\affil{University of Virginia}
TING YAN
\affil{Eaton Innovation Center}
TIAN HE
\affil{University of Minnesota}
CHENGDU HUANG
\affil{Google}
JOHN A. STANKOVIC
\affil{University of Virginia}
TAREK F. ABDELZAHER
\affil{University of Illinois at Urbana-Champaign}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

@@

\category{C.2.2}{Computer-Communication Networks}{Network Protocols}

\terms{Design, Algorithms, Performance}

@@

\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic,
and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
designed for  wireless sensor network applications.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

@@

Author's addresses: G. Zhou, Computer Science Department,
College of William and Mary; Y. Wu  {and} J. A. Stankovic,
Computer Science Department, University of Virginia; T. Yan,
Eaton Innovation Center; T. He, Computer Science Department,
University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
(Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

@@

Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution. The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1, fluidity_manual_v4}), whose sizes depend on the order of the method. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Object name & Type & Variable name(s) \\\hline
Determinant of the Jacobian matrix & double & det  \\ 
Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
Coordinates & double** & coords\\ 
Fields & double** & w \\ 
Numerical integration weights & double[] & W \\ 
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\label{table:map-name-letters}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{C|C}
\hline
Name (Abbreviation) & Parameter \\\hline
Generalized loop-invariant code motion (\emph{licm}) &   \\ 
Padding &  \\ 
Data Alignment & \\ 
Loop interchange      & loops  \\ 
Loop unrolling  & unroll factor \\ 
Outer-product vectorization (\emph{op-vect}) & tile size \\ 
Expression splitting (\emph{split}) & split point, split factor \\ \hline
\end{tabulary}
\end{center}
\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
\label{table:code-transformations}
\end{table}

@@

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
\cline{3-18}
& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
\cline{3-18}
& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
\cline{1-18}
& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1,49 & 1,06 & 1,05 & 1,17 & 1,50 & 2,41 & 1,30 & 1,96 \\
\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1,28 & 1,29 & 2,05 & 1,73 & 1,41 & 1,50 & 2,79 & 2,81 \\
\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 1,04 & 2,26 & 1,93 & 1,64 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
%\cline{1-18}
\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 1,07 & 1,06 & 1,18 & 1,16 & 1,08 & 1,88 & 1,20 & 1,97\\
\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 1,00 & 1,38 & 2,02 & 1,74 & 1,05 & 1,51 & 2,76 & 3,00\\
\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 1,11 & 2,16 & 1,85 & 2,83 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
%\cline{1-18}
\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1,21 & 1,42 & 2,34 & 2,97 & 2,84 & 2,26 & 3,96 & 4,27 \\
\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1,01 & 2,55 & 0,98 & 1,21 & 1,48 & 3,83 & 1,55 & 1,29 \\
\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1,39 & 1,56 & 1,18 & 1,04 & 2,18 & 2,82 & 1,24 & 1,25 \\
\cline{1-18}
\end{tabularx}
}

\begin{table*}[t]\footnotesize
\massmatrixtwodresultsnorms
~
\caption{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code. The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding. Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}
\label{table:perf-results-licm}
\end{table*}

@@

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Problem & Max slow down & Max speed up \\\hline
Helmholtz & - & 4.14$\times$ \\
Diffusion & - & 4.28$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
\end{center}
\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
\label{table:comparison-to-FFC-opt}
\end{table}

@@

% Electronic Appendix
\elecappendix

>>>>

\usepackage{tabulary}
\usepackage[cmex10]{amsmath}
\usepackage{url}

@@

% Title portion
\title{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
\author{Fabio Luporini
\affil{Imperial College London}
Ana Lucia Varbanescu
\affil{University of Amsterdam}
Florian Rathgeber
\affil{Imperial College London}
Gheorghe-Teodor Bercea
\affil{Imperial College London}
J. Ramanujam
\affil{Louisiana State University}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

@@

\category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}

\terms{Design, Performance}

@@

\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber, Gheorghe-Teodor Bercea, J. Ramanujam, David A. Ham, and Paul H. J. Kelly, 2014. COFFEE: an Optimizing Compilerfor Finite Element Local Assembly.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

@@

Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$ Gheorghe-Teodor Bercea $\&$ Paul H. J. Kelly, Department of Computing, Imperial College London; Ana Lucia Varbanescu, Department of Computer Sciences, University of Amsterdam; David A. Ham, Department of Computing and Department of Mathematics, Imperial College London; J. Ramanujam, Department of Electrical and Computer Engineering, Louisiana State University.
\end{bottomstuff}

@@

Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution. The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose sizes depend on the order of the method. This operation is applied to all cells in the discretized domain. In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.

@@

\begin{table}[h]
\begin{center}
\tbl{Type and variable names used in the various code snippets to identify local assembly objects.}{
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Object name & Type & Variable name(s) \\\hline
Determinant of the Jacobian matrix & double & det  \\ 
Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
Coordinates & double** & coords\\ 
Fields & double** & w \\ 
Numerical integration weights & double[] & W \\ 
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
Element matrix & double[][] & A\\ \hline
\end{tabulary}
}
\end{center}
\label{table:map-name-letters}
\end{table}

@@

\begin{table}[h]
\begin{center}
\tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
\begin{tabulary}{1.0\columnwidth}{C|C}
\hline
Name (Abbreviation) & Parameter \\\hline
Generalized loop-invariant code motion (\emph{licm}) &   \\ 
Padding &  \\ 
Data Alignment & \\ 
Loop interchange      & loops  \\ 
Loop unrolling  & unroll factor \\ 
Outer-product vectorization (\emph{op-vect}) & tile size \\ 
Expression splitting (\emph{split}) & split point, split factor \\ \hline
\end{tabulary}
}
\end{center}
\label{table:code-transformations}
\end{table}

@@

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabulary}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
\cline{3-18}
& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
\cline{3-18}
& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
\cline{1-18}
& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1,49 & 1,06 & 1,05 & 1,17 & 1,50 & 2,41 & 1,30 & 1,96 \\
\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1,28 & 1,29 & 2,05 & 1,73 & 1,41 & 1,50 & 2,79 & 2,81 \\
\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 1,04 & 2,26 & 1,93 & 1,64 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
%\cline{1-18}
\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 1,07 & 1,06 & 1,18 & 1,16 & 1,08 & 1,88 & 1,20 & 1,97\\
\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 1,00 & 1,38 & 2,02 & 1,74 & 1,05 & 1,51 & 2,76 & 3,00\\
\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 1,11 & 2,16 & 1,85 & 2,83 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
%\cline{1-18}
\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1,21 & 1,42 & 2,34 & 2,97 & 2,84 & 2,26 & 3,96 & 4,27 \\
\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1,01 & 2,55 & 0,98 & 1,21 & 1,48 & 3,83 & 1,55 & 1,29 \\
\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1,39 & 1,56 & 1,18 & 1,04 & 2,18 & 2,82 & 1,24 & 1,25 \\
\cline{1-18}
\end{tabulary}
}

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code. The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding. Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}{
\scriptsize
\massmatrixtwodresultsnorms
}
\label{table:perf-results-licm}
\end{table*}

@@

\begin{table}[h]
\begin{center}
\tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}{
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Problem & Max slow down & Max speed up \\\hline
Helmholtz & - & 4.14$\times$ \\
Diffusion & - & 4.28$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
}
\end{center}
\label{table:comparison-to-FFC-opt}
\end{table}

@@
